---
title: "Understanding mitochondrial changes in hypoxic cell conditions"
output:
  html_document: 
    df_print: kable
    css: styles.css
    highlight: tango
    theme: yeti
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("C:/Users/Aditi/Documents/Desktop/R/DGE Analysis Main")
```

# Loading libraries

```{r message=FALSE, warning=FALSE}
library(DESeq2)
library(ReactomePA)
library(enrichplot)
library(tidyverse)
library(RColorBrewer)
library(pheatmap)
library(DEGreport)
library(tximport)
library(ggplot2)
library(ggrepel)
library(apeglm)
library(reshape2)
library(edgeR)
library(AnnotationHub)
library(ensembldb)
library(pathview)
library(clusterProfiler)
library(org.Hs.eg.db)
library(conflicted)
library(ensembldb)
library(dplyr)
###### ANNOTATION
# Connect to AnnotationHub
ah <- AnnotationHub()
human_ens <- query(ah, c("Homo sapiens", "EnsDb"))
human_ens
human_ens <- human_ens[["AH116860"]]
```

# Inspecting the data

First, load the raw counts data.

```{r echo=TRUE, message=FALSE, warning=FALSE}
#load data
raw_counts <- read.table("./data/GSE132624/GSE132624_raw_counts.tsv", header = TRUE, sep = "\t", row.names = 1)
View(raw_counts)
tail(raw_counts)
```

```{r}
# Create vectors for each column
sample_id <- c(
  paste0("GSM38823", 37:48),
  paste0("GSM38823", 49:60),
  paste0("GSM38823", 61:66),
  paste0("GSM38823", c(67, 69)),  # Skip 68 for 24h
  paste0("GSM38823", 70:72)
)

condition <- c(rep("501", 12), rep("IGR37", 12), rep("IGR39", 11))

time <- c(
  rep(c(0, 12, 24, 48), each = 3),  # For 501
  rep(c(0, 12, 24, 48), each = 3),  # For IGR37
  rep(c(0, 12), each = 3),          # For IGR39 (0h and 12h)
  rep(24, 2),                       # For IGR39 (24h, only 2 reps)
  rep(48, 3)                        # For IGR39 (48h)
)

replicate <- rep(c(1, 2, 3), length.out = length(sample_id))

# Create the treatment column
treatment <- ifelse(time == 0, "normoxia", "hypoxia")

# Create the data frame
meta <- data.frame(
  condition = condition,
  time = time,
  replicate = replicate,
  treatment = treatment,
  row.names = sample_id
)

# Write the metadata to a CSV file
write.csv(meta, file="./data/meta_132624.csv")
# Display the full metadata
print(meta)
```

Let's look at the distribution of raw counts in one of the samples.

```{r message=FALSE, warning=FALSE}
#distribution of counts in samples
ggplot(raw_counts) +
  geom_histogram(aes(x = GSM3882337), stat = "bin", bins = 200) +
  xlim(-5,1000) +
  xlab("Raw expression counts") +
  ylab("Number of genes")
```

Filter genes with low counts (CPM) for higher DGE accuracy.

```{r message=FALSE, warning=FALSE}
  #Density Plot of the log-transformed Raw Count data 
  df_log2<-log2(raw_counts)
  df_log2_melted<-melt(df_log2)
Raw_Count_Density_Plot <- ggplot(df_log2_melted, aes(x=value, fill=variable)) +
    geom_density(alpha=.25) +
    labs(x='Log2 Counts', y='Density', title='Raw Count Distribution (Before Filtering)') +
    guides(fill = FALSE)  # Suppress the legend entirely

Raw_Count_Density_Plot
  
  # Create a DGEList object
  dge <- DGEList(counts = raw_counts)

  # Calculate library sizes
  dge$samples$lib.size <- colSums(dge$counts)
  
  # Calculate CPM
  cpm_counts <- cpm(dge)
  
  #Plot CPM vs Raw Counts (Run the 3 commands together)
  plot(cpm_counts[,1],raw_counts[,1],ylim=c(0,50),xlim=c(0,10),xlab="CPM", ylab="Raw Counts")
  abline(v=1.9)
  abline(h=10)
  
  # Set your CPM threshold (e.g., the value of CPM on the x axis which intersects with a Raw Count value of ~ 10)
  cpm_threshold <- 1.9
  
  # Set the minimum number of samples that should meet the CPM threshold
  min_samples <- 2
  
  # Create a logical matrix indicating which genes meet the CPM threshold in each sample
  genes_to_keep <- rowSums(cpm_counts >= cpm_threshold) >= min_samples
  
  # Filter the raw counts matrix
  counts_filtered <- raw_counts[genes_to_keep, ]

  # Print some statistics
  cat("Number of genes before filtering:", nrow(raw_counts), "\n")
  cat("Number of genes after filtering:", nrow(counts_filtered), "\n")
  cat("Percentage of genes retained:", round(nrow(counts_filtered) / nrow(raw_counts) * 100, 2), "%\n")
  
  # Write the filtered counts to a new file
  # write.table(counts_filtered, "counts_filtered.txt", sep = "\t", quote = FALSE)
  
  #Plotting Sample Density Distributions after low count filtering
  df_log2<-log2(counts_filtered)
  df_log2_melted<-melt(df_log2)
Filtered_Count_Density_Plot <- ggplot(df_log2_melted, aes(x=value, fill=variable)) +
    geom_density(alpha=.25) +
    labs(x='Log2 Intensity', y='Density', title='Raw Count Distribution (After Filtering)') +
    guides(fill = FALSE)  # Suppress the legend entirely

Filtered_Count_Density_Plot
  
# Save the density plot before filtering
ggsave("Raw_Count_Density_Plot.png", plot = Raw_Count_Density_Plot, width = 8, height = 6, dpi = 300)

# Save the density plot after filtering
ggsave("Filtered_Count_Density_Plot.png", plot = Filtered_Count_Density_Plot, width = 8, height = 6, dpi = 300)

```

Next, lets look at mean vs. variance to inspect the data further.

```{r message=FALSE, warning=FALSE}
#mean vs variance
mean_counts <- apply(counts_filtered[, 1:3], 1, mean)
variance_counts <- apply(counts_filtered[, 1:3], 1, var)
df <- data.frame(mean_counts, variance_counts)
  
ggplot(df) +
  geom_point(aes(x=mean_counts, y=variance_counts)) + 
  geom_line(aes(x=mean_counts, y=mean_counts, color="red")) +
  scale_y_log10() +
  scale_x_log10()
mean_counts <- apply(counts_filtered[, 21:23], 1, mean)
variance_counts <- apply(counts_filtered[, 21:23], 1, var)
df <- data.frame(mean_counts, variance_counts)

ggplot(df) +
  geom_point(aes(x=mean_counts, y=variance_counts)) + 
  geom_line(aes(x=mean_counts, y=mean_counts, color="red")) +
  scale_y_log10() +
  scale_x_log10()
mean_counts <- apply(counts_filtered[, 33:35], 1, mean)
variance_counts <- apply(counts_filtered[, 33:35], 1, var)
df <- data.frame(mean_counts, variance_counts)

ggplot(df) +
  geom_point(aes(x=mean_counts, y=variance_counts)) + 
  geom_line(aes(x=mean_counts, y=mean_counts, color="red")) +
  scale_y_log10() +
  scale_x_log10()
```

------------------------------------------------------------------------

In all 3 cases, the mean ≠ variance and there is noticeable heteroscedasticity among genes with lower expression in all 3 groups. Additionally, variance \> mean for genes with high mean expression.

Create the metadata file containing sample information.

```{r}
# Check that sample names match in both files
all(colnames(counts_filtered) %in% rownames(meta))
```

Now, create the DESeq2 object with `raw_counts` and `meta` as inputs.

```{r message=FALSE, warning=FALSE}
# Create the DESeqDataSet
meta$time <- factor(meta$time)
dds <- DESeqDataSetFromMatrix(countData = counts_filtered, colData = meta, design = ~ condition + replicate + time)
View(counts(dds))
```

We will generate a normalized counts table from dds to be used for data visualization before actually running differential gene expression analysis using DESeq2.

```{r message = FALSE, warning = FALSE}
dds <- estimateSizeFactors(dds)
sizeFactors(dds)
normalized_counts <- counts(dds, normalized=TRUE)
write.table(normalized_counts, file="./data/normalized_counts_132624.txt", sep="\t", quote=F, col.names=NA)
```

Log transformation is applied on the normalized counts to perform dimensionality reduction (principal component analysis) and hierarchical clustering.

```{r message=FALSE, warning=FALSE}
vsd <- vst(dds, blind = T)

pca_data <- plotPCA(vsd, intgroup = c("condition","time"), returnData = T)
ggplot(pca_data, aes(PC1, PC2, color = group)) +
  geom_point(size = 3) +
  labs(x = paste0("PC1: ", round(100 * attr(pca_data, "percentVar")[1], 2), "%"),
       y = paste0("PC2: ", round(100 * attr(pca_data, "percentVar")[2], 2), "%"),
       title = "PCA Plot") +
  theme_minimal()
ntop <- 500
expr_matrix <- assay(vsd)
rv <- rowVars(expr_matrix)
select <- order(rv, decreasing = TRUE)[seq_len(min(ntop, length(rv)))]
expr_matrix_top <- expr_matrix[select, ]

# Perform PCA on the top 500 genes
pca<- prcomp(t(expr_matrix_top), center = TRUE, scale. = TRUE)

# c5alculate variance and cumulative variance
eigen_values <- tibble(
  PC = factor(1:length(pca$sdev)),
  var_pct = (pca$sdev^2 / sum(pca$sdev^2)) * 100,
  cum_var = cumsum((pca$sdev^2 / sum(pca$sdev^2)) * 100)
)

# scree plot
ggplot(eigen_values, aes(x = PC)) +
  geom_col(aes(y = var_pct), fill = "skyblue") +
  geom_line(aes(y = cum_var, group = 1), color = "red") +
  geom_point(aes(y = cum_var), color = "red") +
  labs(title = "Scree Plot: top 500", x = "Principal Components", y = "Variance Explained (%)") +
  theme_minimal()

write.table(expr_matrix, file="./data/expr_matrix_132624.txt", sep="\t", quote=F, col.names=NA)
```

Extract PC loadings.

```{r message=FALSE, warning=FALSE}
# Extract loadings for PC1
pc1_loadings <- pca$rotation[, 1]
pc2_loadings <- pca$rotation[, 2]

# Sort genes by their PC1 loadings
sorted_genes_1 <- sort(pc1_loadings, decreasing = TRUE)
sorted_genes_2 <- sort(pc2_loadings, decreasing = TRUE)

# Extract top 100 positive and negative genes
top_100_positive_1 <- data.frame(head(sorted_genes_1, 100))
top_100_negative_1 <- data.frame(tail(sorted_genes_1, 100))
top_100_positive_2 <- data.frame(head(sorted_genes_2, 100))
top_100_negative_2 <- data.frame(tail(sorted_genes_2, 100))

pca_rawcounts_1 <- counts_filtered
pca_rawcounts_1 <- rownames_to_column(pca_rawcounts_1, var = "entrezid")

pca_rawcounts_2 <- counts_filtered
pca_rawcounts_2 <- rownames_to_column(pca_rawcounts_2, var = "entrezid")

annotations_ahb <- genes(human_ens, return.type = "data.frame") %>%
  dplyr::select(gene_name, gene_id, entrezid, gene_biotype, description) %>%
  dplyr::filter(entrezid %in% pca_rawcounts_1$entrezid) %>%
  dplyr::distinct(entrezid, .keep_all = TRUE)

annotations_ahb$entrezid <- as.character(unlist(annotations_ahb$entrezid))

top_100_positive_1 <- rownames_to_column(top_100_positive_1, var = "entrezid")
top_100_positive_1 <- left_join(top_100_positive_1, annotations_ahb, by = "entrezid")

top_100_positive_2 <- rownames_to_column(top_100_positive_2, var = "entrezid")
top_100_positive_2 <- left_join(top_100_positive_2, annotations_ahb, by = "entrezid")

top_100_negative_1 <- rownames_to_column(top_100_negative_1, var = "entrezid")
top_100_negative_1 <- left_join(top_100_negative_1, annotations_ahb, by = "entrezid")

top_100_negative_2 <- rownames_to_column(top_100_negative_2, var = "entrezid")
top_100_negative_2 <- left_join(top_100_negative_2, annotations_ahb, by = "entrezid")

head(top_100_positive_1[, c("entrezid", "gene_name", "gene_biotype", "description")], n = 20)
head(top_100_negative_1[, c("entrezid","gene_name", "gene_biotype", "description")], n = 20)
head(top_100_positive_2[, c("entrezid", "gene_name", "gene_biotype", "description")], n = 20)
head(top_100_negative_2[, c("entrezid","gene_name", "gene_biotype", "description")], n = 20)
```

```{r message=FALSE, warning=FALSE}
# Compute pairwise correlation values
vsd_cor <- cor(expr_matrix)
head(vsd_cor)
# Plot heatmap
plot1 <- pheatmap(vsd_cor, annotation = meta, angle_col = 45, show_colnames = F, show_rownames = F)
ggsave("corr heat map.png", plot = plot1, width = 10, height = 8, dpi = 300)
```

# Likelihood Ratio Test

```{r message=FALSE, warning=FALSE}
#Likelihood ratio test
dds_lrt <- DESeq(dds, test="LRT", reduced = ~ condition + replicate)
# Extract results
res_LRT <- results(dds_lrt)
# Create a tibble for LRT results
res_LRT_tb <- res_LRT %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>% 
  as_tibble()

# Subset to return genes with padj < 0.01
sigLRT_genes <- res_LRT_tb %>% 
  dplyr::filter(padj < 0.01)

# Obtain vst values for those significant genes
cluster_vst <- expr_matrix[sigLRT_genes$gene, ]
```

```{r deg-patterns-plot, fig.width=16, fig.height=10, message = FALSE, warning = FALSE}
clusters <- degPatterns(cluster_vst, metadata = meta, time = "time", col = "condition")
ggsave("degPatterns_Plot.png", plot = clusters$plot, width = 20, height = 18, dpi = 300)
```

```{r message=FALSE, warning=FALSE}
# Define the fix function
fix <- function(x) {
  x$gene <- rownames(x)
  x$genes <- NULL
  rownames(x) <- NULL
  return(x)
}
rownames(clusters$df) <- sub("^X", "", rownames(clusters$df))
# Initialize an empty list to store the groups
group_list <- list()

for (i in 1:45) {
  group_list[[paste0("group", i)]] <- clusters$df %>%
    dplyr::filter(cluster == i)
    group_list[[paste0("group", i)]] <- fix(group_list[[paste0("group", i)]])
}
group_list <- group_list[sapply(group_list, function(x) nrow(x) > 0)]
for (name in names(group_list)) {
  cluster_num <- unique(group_list[[name]]$cluster)  # Get the cluster number
  file_name <- paste0("cluster", cluster_num, ".csv")  # Create the correct file name
  write.csv(group_list[[name]], file_name, row.names = FALSE)  # Save as CSV
}
```

------------------------------------------------------------------------

# Functional Analysis

Annotation hub is used here to obtain gene information such as their entrez ids, their biotype, gene symboles, and enzembl ids. (e.g. protein coding, lncRNA, miRNA, etc.)

```{r}
# Initialize an empty list to store enrichment results
enrichment_results <- list()

# Loop through the group_list and perform pathway enrichment
for (name in names(group_list)) {
  genes <- group_list[[name]]$gene
  cluster_num <- unique(group_list[[name]]$cluster)  # Get the actual cluster number

  enrichment <- enrichPathway(gene = genes,
                              organism = "human",
                              pvalueCutoff = 0.01,
                              pAdjustMethod = "BH",
                              readable = TRUE)
  
  if (nrow(enrichment@result) > 0) {
    enrichment_results[[paste0("cluster", cluster_num)]] <- enrichment
    # Write the enrichment results to a CSV file using the correct cluster number
    write.csv(enrichment@result, file = paste0("enrichment_results_cluster", cluster_num, ".csv"), row.names = FALSE)
  }
}

# Create dotplots for each cluster with significant results
for (name in names(enrichment_results)) {
  if (length(enrichment_results[[name]]) > 0 && nrow(enrichment_results[[name]]@result) > 0) {
    # Filter for significant results
    significant_results <- enrichment_results[[name]]@result %>% 
      filter(p.adjust < 0.01)
    
    if (nrow(significant_results) > 0) {
      p <- dotplot(enrichment_results[[name]], showCategory = 12, label_format = 70) +
        ggtitle(paste("Enrichment Dotplot for", name))
      
      # Save dotplot for the corresponding cluster
      ggsave(paste0("enrichment_dotplot_", name, ".png"), plot = p, width = 12, height = 7, dpi = 300)
    } else {
      message(paste("No significant pathways (p.adjust < 0.01) found for", name))
    }
  } else {
    message(paste("No enrichment results found for", name))
  }
}

# Create a combined dotplot for all clusters
all_results <- do.call(rbind, lapply(names(enrichment_results), function(name) {
  result <- enrichment_results[[name]]@result
  if (nrow(result) > 0) {
    significant_results <- result %>% filter(p.adjust < 0.01)
    if (nrow(significant_results) > 0) {
      significant_results$Cluster <- name  # Use the actual cluster name
      return(significant_results)
    }
  }
  return(NULL)
}))

# If significant pathways exist, create a combined dotplot
if (!is.null(all_results) && nrow(all_results) > 0) {
  all_results <- all_results[order(all_results$p.adjust), ]
  top_pathways <- all_results %>%
    group_by(Cluster) %>%
    slice_head(n = 5) %>%
    ungroup()

  p_combined <- ggplot(top_pathways, aes(x = Cluster, y = Description, size = Count, color = p.adjust)) +
    geom_point() +
    scale_color_gradient(low = "red", high = "blue") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(x = "Cluster", y = "Pathway", size = "Gene Count", color = "Adjusted p-value") +
    ggtitle("Top 5 Enriched Pathways Across All Clusters")

  ggsave("combined_enrichment_dotplot.png", plot = p_combined, width = 20, height = 12, dpi = 300)
} else {
  message("No significant enrichment results (p.adjust < 0.01) found across all clusters")
}
```

# Subset Mitochondrial genes

Repeat the same process for a subset of genes corresponding to mitochondria from the significant genes obtained from the Likelihood Ratio Test.

```{r deg-patterns-mit, fig.width=16, fig.height=10, message = FALSE, warning = FALSE}
#subset mitochondrial genes
mito_genes <- read.table("C:/Users/Aditi/Documents/Desktop/R/DGE Analysis Main/data/HumanMitoCarta.txt", header = TRUE, sep = "\t")
entrez_ids <- mito_genes$HumanGeneID
sigLRT_genes_mit <- sigLRT_genes %>% filter(sigLRT_genes$gene %in% entrez_ids)

cluster_vst <- expr_matrix[sigLRT_genes_mit$gene, ]
clusters <- degPatterns(cluster_vst, metadata = meta, time = "time", col = "condition")
# Save the plot generated by degPatterns
ggsave("degPatterns_Plot_mit.png", plot = clusters$plot, width = 20, height = 18, dpi = 300)

# Define the fix function
fix <- function(x) {
  x$gene <- rownames(x)
  x$genes <- NULL
  rownames(x) <- NULL
  return(x)
}
rownames(clusters$df) <- sub("^X", "", rownames(clusters$df))
# Initialize an empty list to store the groups
group_list <- list()

for (i in 1:45) {
  group_list[[paste0("group", i)]] <- clusters$df %>%
    dplyr::filter(cluster == i)
    group_list[[paste0("group", i)]] <- fix(group_list[[paste0("group", i)]])
}
group_list <- group_list[sapply(group_list, function(x) nrow(x) > 0)]
for (name in names(group_list)) {
  cluster_num <- unique(group_list[[name]]$cluster)  # Get the cluster number
  file_name <- paste0("cluster_mit", cluster_num, ".csv")  # Create the correct file name
  write.csv(group_list[[name]], file_name, row.names = FALSE)  # Save as CSV
}
# Initialize an empty list to store enrichment results
enrichment_results <- list()

# Perform pathway enrichment and save results
for (name in names(group_list)) {
  genes <- group_list[[name]]$gene
  cluster_num <- unique(group_list[[name]]$cluster)  # Get actual cluster number

  enrichment <- enrichPathway(gene = genes,
                              organism = "human",
                              pvalueCutoff = 0.01,
                              pAdjustMethod = "BH",
                              readable = TRUE)

  if (nrow(enrichment@result) > 0) {
    enrichment_results[[paste0("cluster", cluster_num)]] <- enrichment
    write.csv(enrichment@result, file = paste0("enrichment_results_cluster", cluster_num, ".csv"), row.names = FALSE)
  }
}

# Create dotplots for each cluster
for (name in names(enrichment_results)) {
  significant_results <- enrichment_results[[name]]@result %>% filter(p.adjust < 0.01)
  
  if (nrow(significant_results) > 0) {
    p <- dotplot(enrichment_results[[name]], showCategory = 12, label_format = 70) +
         ggtitle(paste("Enrichment Dotplot for", name))
    ggsave(paste0("enrichment_dotplot_", name, ".png"), plot = p, width = 12, height = 7, dpi = 300)
  } else {
    message(paste("No significant pathways (p.adjust < 0.01) found for", name))
  }
}

# Create a combined dotplot for all clusters
all_results <- do.call(rbind, lapply(names(enrichment_results), function(name) {
  enrichment_results[[name]]@result %>% filter(p.adjust < 0.01) %>% 
    mutate(Cluster = name)
}))

if (!is.null(all_results) && nrow(all_results) > 0) {
  top_pathways <- all_results %>% group_by(Cluster) %>% slice_head(n = 5)

  p_combined <- ggplot(top_pathways, aes(x = Cluster, y = Description, size = Count, color = p.adjust)) +
                geom_point() +
                scale_color_gradient(low = "red", high = "blue") +
                theme_minimal() +
                theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
                labs(x = "Cluster", y = "Pathway", size = "Gene Count", color = "Adjusted p-value") +
                ggtitle("Top 5 Enriched Pathways Across All Clusters")

  ggsave("combined_enrichment_dotplot.png", plot = p_combined, width = 20, height = 12, dpi = 300)
} else {
  message("No significant enrichment results (p.adjust < 0.01) found across all clusters")
}
```
